On Verification Patterns for Low-Level Systems via Modal Abstractions

### 2025-09-11 paper.tex
+

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_{\ell}\varphi$ asserts that $\varphi$ is true in the state identified by $\ell$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $\ell$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.

@@

In each of these cases, the fact that these facts are described using modalities with the core modal property $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$ means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion \[ \mathsf{minCommitIndex}(5)\] to represent the former knowledge,
one would need to separately provide custom reasoning to conclude \[ \mathsf{minCommitIndex}(3)\].
As our survey shows, the value of the modal view (which grants both intuitive specifications and natural
reasoning principles) is becoming apparent in systems software verification.

>>>>

\section{Background on Modal Logics for Programs, and Low-Level Systems Software}

This section briefly recalls related concepts from modal logic of particular relevance
to our taxonomy, which should be familiar but benefit from fixing notation (Section \ref{sec:modalbackground}),
and enough general background on the systems software concepts organized by our taxonomy
for the connections to modalities to be clear (Section \ref{sec:background}).

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_{\ell}\varphi$ asserts that $\varphi$ is true in the state identified by $\ell$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.

@@

In each of these cases, the fact that these facts are described using modalities with the core modal property $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$ means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion \[\mathsf{minCommitIndex}(5) \] to represent the former knowledge,
one would need to separately provide custom reasoning to conclude \[\mathsf{minCommitIndex}(3)\].
As our survey shows, the value of the modal view (which grants both intuitive specifications and natural
reasoning principles) is becoming apparent in systems software verification.
